A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-23 sections/Overview.tex
Section 2.5

Our new algorithmic type system addresses builds on the work above, but
addresses the open problems. 
\bruno{I think this section needs more detail here. The textual
  description alone does not convey much about the key ideas of the
  approach. I suggest showing the forms of the judgements and maybe
  illustrating with some examples why the blend between ordered
  contexts and worklists leads to precise scope traking. This can be
  done just by illustrating a concrete instance of an algorithmic
  context (no derivation is actually needed). Similar things should be
done for inference judgements. Actually chains needs some careful
explanation, I think since they are the hardest to get aspect of the formalization.}

\jimmy{Some high level description for the structure}
A worklist is structurally a stack with its top element to the right.
$$\begin{aligned}
\text{Algorithmic judgement chain}\qquad&\jg &::=&\quad A \le B \mid e\Lto A \mid e\To_{a} \jg \mid \appInfAlg{A}{e}\\
\text{Algorithmic worklist}\qquad&\Gm &::=&\quad \nil \mid \Gm, a \mid \Gm, \al \mid \Gm, x: A \mid \Gm \Vdash \jg\\
\end{aligned}$$

\paragraph{Scope Tracking}
Firstly, we avoid scoping garbage by blending the ordered context and the
worklist into a single syntactic sort $\Gamma$, our algorithmic worklist. This
algorithmic worklist interleaves (type and term) variables with \emph{work}
like checking or inferring types of expressions. The interleaving keeps track
of the variable scopes in the usual, natural way: each variables is in scope of
anything in front of it in the worklist. If there is nothing in front, the
variable is no longer needed and can be popped from the worklist. This way, no
garbage builds up.

For example, the top judgment of the following worklist
checks the identity function against $\all a \to a$:
$$\Gm \Vdash \lam x \Lto \all a \to a$$
Then, two auxiliary variables $a$ and $x$ are introduced to help the type checking:
$$\Gm, a, x:a \Vdash x \Lto a$$
which will be satisfied after several steps, and the worklist becomes
$$\Gm, a, x:a$$
Since the variable declarations $a, x:a$ are only used for a judgment already processed,
they can be safely removed, leaving the remaining worklist $\Gm$ to be further reduced.

@@

\paragraph{Inference Judgements}
To express the DK inference judgements, we use a novel form of work entries in
the worklist: our algorithmic judgement chains $\omega$. In its simplest form,
such a judgement chain is just a check, like a subtyping check $A \leq B$ or a
term typecheck $e \Leftarrow A$.  The non-trivial forms of chains capture an
inference task together with the the work that depends on its outcome. For
instance, a type inference task takes the form $e \Rightarrow_a \omega$. It
denotes that we need to infer the type, say $A$, of expression $e$ and use it
in the chain $\omega$ by subsituting it for the placeholder type variable $a$.

@@

The algorithm first analyse the two subtyping judgments and
will find the best solutions $\al = \bt = 1$
(please refer to Figure~\ref{fig:alg} for detailed derivations).
Then we substitute every instance of $\al$ and $\bt$ to $1$,
so the variable declarations can be safely removed from the worklist.
Now we reduce the worklist to the following state
$$\nil \Vdash \blue{(\lam x)~() \To_a a \le 1}, x:1$$
which has a term variable declaration as the top element.
Because no judgments before the declaration refer to $x$,
as a basic property of an ordered context
(Figure~\ref{fig:alg:syntax} shows its well-formedness criteria),
we remove all declarations before the first judgment chain to collect unused variables.
As a result all garbage variables are collected
once the judgment chains refering to them are processed.

After that, we proceed the last judgment chain from the left,
namely an inference judgment $(\lam x)~()\To~?$, with the unit type $1$ as its result.
Finally the last judgment becomes $1 \le 1$, and is satisfied as a base case.

>>>>

Our new algorithmic type system addresses builds on the work above, but
addresses the open problems.

\paragraph{Scope Tracking}
We avoid scoping garbage by blending the ordered context and the
worklist into a single syntactic sort $\Gamma$, our algorithmic worklist. This
algorithmic worklist interleaves (type and term) variables with \emph{work}
like checking or inferring types of expressions. The interleaving keeps track
of the variable scopes in the usual, natural way: each variable is in scope of
anything in front of it in the worklist. If there is nothing in front, the
variable is no longer needed and can be popped from the worklist. This way, no
garbage (i.e. variables out-of-scope) builds up.

$$\begin{aligned}
\text{Algorithmic judgement chain}\qquad&\jg &::=&\quad A \le B \mid e\Lto A \mid e\To_{a} \jg \mid \appInfAlg{A}{e}\\
\text{Algorithmic worklist}\qquad&\Gm &::=&\quad \nil \mid \Gm, a \mid \Gm, \al \mid \Gm, x: A \mid \Gm \Vdash \jg\\
\end{aligned}$$

For example, suppose that the top judgment of the following worklist
checks the identity function against $\all a \to a$:
$$\Gm \Vdash \lam x \Lto \all a \to a$$
To proceed, two auxiliary variables $a$ and $x$ are introduced to help the type checking:
$$\Gm, a, x:a \Vdash x \Lto a$$
which will be satisfied after several steps, and the worklist becomes
$$\Gm, a, x:a$$
Since the variable declarations $a, x:a$ are only used for a judgment already processed,
they can be safely removed, leaving the remaining worklist $\Gm$ to be further reduced.

Our worklist can be seen as an all-in-one stack,
containing variable declarations and subtyping/typing judgments.
The stack is an enriched form of ordered context,
and it has a similar variable scoping scheme.
%The well-formedness condition, shown in Figure~\ref{fig:alg:syntax},
%states that every variable must be declared prior to it usage.
%Therefore, if any variable declaration is the top element,
%then no usage should appear in the rest of the worklist, so it can be removed.

@@

\paragraph{Inference Judgements}
To express the DK's inference judgements, we use a novel form of work entries in
the worklist: our algorithmic judgement chains $\omega$. In its simplest form,
such a judgement chain is just a check, like a subtyping check $A \leq B$ or a
term typecheck $e \Leftarrow A$. 
However, the non-trivial forms of chains capture an
inference task together with the the work that depends on its outcome. For
instance, a type inference task takes the form $e \Rightarrow_a \omega$. It
denotes that we need to infer the type, say $A$, for expression $e$ and use it
in the chain $\omega$ by substituting it for the placeholder type variable $a$.

@@

The algorithm first analyses the two subtyping judgments and
will find the best solutions $\al = \bt = 1$
(please refer to Figure~\ref{fig:alg} for detailed derivations).
Then we substitute every instance of $\al$ and $\bt$ to $1$,
so the variable declarations can be safely removed from the worklist.
Now we reduce the worklist to the following state
$$\nil \Vdash \blue{(\lam x)~() \To_a a \le 1}, x:1$$
which has a term variable declaration as the top element.
%Because no judgments before the declaration refer to $x$,
%as a basic property of an ordered context
%(Figure~\ref{fig:alg:syntax} shows its well-formedness criteria),
%we remove all declarations before the first judgment chain to collect unused variables.
%As a result all garbage variables are collected
%once the judgment chains refering to them are processed,
%just as what we have mentioned above about the scope tracking.
After removing the garbage term variable declaration from the worklist, we process the last judgment chain from the left,
namely an inference judgment $(\lam x)~()\To~?$, with the unit type $1$ as its result.
Finally the last judgment becomes $1 \le 1$, and is satisfied as a base case.
